Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-12-24 paper/paper.tex
removed Steuerman reanalysis (have too many supp figs and need to reduce)

A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
Studies using reporters~\citep{killip2017single} and our own previous single-cell transcriptomics~\citep{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
IFN expression also appears to be rare among cells in influenza-infected mice~\citep{kallfass2013visualizing}, a conclusion supported by our re-analysis of single-cell transcriptomics of influenza-infected mice by Steuerman et al~\citep{steuerman2018dissection} (\SUPPFIG{mice}).

To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\citep{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

@@

It turns out that these viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\citep{velthuis2018mini} .
Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
Given the extensive prior work describing such deletions, it is perhaps surprising we did not identify more of them in our single IFN+ cells.
There may be several reasons: we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
The relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.

@@

\begin{suppfig}
\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}
\caption{
{\bf Re-analysis of a published single-cell transcriptomic study~\citep{steuerman2018dissection} of influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.}
Specifically, we downloaded the data from \citep{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \citep{steuerman2018dissection}.
Here we show statistics for the cells from the influenza-infected wild-type C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
As shown in the left-most panel, the sequencing depth of \citep{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in the current study).
An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
Nonetheless, there is detectable expression of key ISGs---which are induced by autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}---in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, which are only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.
The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
}
\label{suppfig:mice}
\end{suppfig}

>>>>

A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
Studies using reporters~\citep{killip2017single,kallfass2013visualizing} and our own previous single-cell transcriptomics~\citep{russell2018extreme} have found that IFN induction is rare among influenza-infected cells, at least early in infection prior to extensive paracrine signaling and the accumulation of defective viral particles.
Therefore, to efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\citep{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
For the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

@@

It turns out that these viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\citep{velthuis2018mini}.
Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
Given the extensive prior work describing such deletions, it is perhaps surprising we did not identify more of them in our single IFN+ cells.
There may be several reasons: we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
The relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.

@@
